---
title: "Final Code Customer Retention"
author: "Annika Rodriguez"
date: "2026-04-23"
output: html_document
---

ANNIKA

```{r}
# Libraries
library(tidyverse)
library(caret)

# Load dataset

df <- read.csv("clean_customerretention_dataset.csv")

df$acquisition <- as.factor(df$acquisition)
df$industry    <- as.factor(df$industry)

# Check structure
str(df)
summary(df)
```



```{r}
#Identify Target Variable
table(df$acquisition)
prop.table(table(df$acquisition))
```



```{r}
colnames(df)
head(df)
```


```{r}
#train/test split

set.seed(123)

train_index <- createDataPartition(df$acquisition, p = 0.80, list = FALSE)

train_data <- df[train_index, ]
test_data  <- df[-train_index, ]

```



```{r}
#EDA visial ex 1

ggplot(df, aes(x = acquisition)) +
  geom_bar(fill = "lightpink") +
  labs(title = "Distribution of Customer Acquisition",
       x = "Acquisition (0 = No, 1 = Yes)",
       y = "Number of Customers") +
  theme_minimal()
```



```{r}
#EDA Visial ex 2

ggplot(df, aes(x = acquisition, y = profit)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Profit by Acquisition Outcome",
       x = "Acquisition",
       y = "Profit") +
  theme_minimal()
```


```{r}
#logistic regression

log_model <- glm(acquisition ~ ., data = train_data, family = "binomial")

summary(log_model)
```


```{r}
#predictions and evals

log_probs <- predict(log_model, newdata = test_data, type = "response")

log_pred <- ifelse(log_probs > 0.5, 1, 0)
log_pred <- as.factor(log_pred)

log_pred <- factor(log_pred, levels = levels(test_data$acquisition))

confusionMatrix(log_pred, test_data$acquisition)
```



```{r}
#removing squared terms to reduce redundancy
log_mods <- glm(acquisition ~ . - acq_exp_sq - ret_exp_sq - freq_sq, data = train_data, family = "binomial")
summary(log_mods)

```


```{r}
#predictions and evals

log_prob <- predict(log_mods, newdata = test_data, type = "response")

log_preds <- ifelse(log_prob > 0.5, 1, 0)
log_preds <- as.factor(log_preds)

log_preds <- factor(log_preds, levels = levels(test_data$acquisition))

confusionMatrix(log_preds, test_data$acquisition)
```



YASMINE


```{r}
# Libraries
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)

# Load dataset
df <- read.csv("clean_customerretention_dataset.csv")

# Check structure
str(df)
summary(df)
```

#Identify Target Variable
```{r}
table(df$acquisition)        # or whatever the column is called
prop.table(table(df$acquisition))
```
```{r}
colnames(df)
head(df)
```

#Step 3: EDA

#1. Target Distribution
```{r}
ggplot(df, aes(x = factor(acquisition))) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Customer Acquisition Distribution",
       x = "Acquired (0 = No, 1 = Yes)",
       y = "Count")
```
#2. Numeric Distributions
```{r}
num_vars <- df %>% select(where(is.numeric))

num_vars |>
  gather() |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "darkgreen") +
  facet_wrap(~key, scales = "free")
```
#3. Boxplots vs Target 
```{r}
num_vars |>
  mutate(acquired = df$acquisition) |>
  gather(key, value, -acquisition) |>
  ggplot(aes(x = factor(acquisition), y = value, fill = factor(acquisition))) +
  geom_boxplot() +
  facet_wrap(~key, scales = "free") +
  theme(legend.position = "none")
```
#4. Correlation Heatmap
```{r}
corr_matrix <- cor(num_vars, use = "complete.obs")

corrplot(corr_matrix, method = "color", type = "upper")
```
#Step 4: Train/Test Split
```{r}
set.seed(123)

df$acquisition <- as.factor(df$acquisition)

trainIndex <- createDataPartition(df$acquisition, p = 0.8, list = FALSE)

train_data <- df[trainIndex, ]
test_data  <- df[-trainIndex, ]

dim(train_data)
dim(test_data)
prop.table(table(train_data$acquisition))
prop.table(table(test_data$acquisition))
```
#Step 5: Decision Tree
```{r}
tree_model <- rpart(acquisition ~ ., 
                    data = train_data, 
                    method = "class",
                    control = rpart.control(cp = 0.01))
```

```{r}
rpart.plot(tree_model)
```

```{r}
pred_tree <- predict(tree_model, test_data, type = "class")
```

```{r}
confusionMatrix(pred_tree, test_data$acquisition)
```
#How strong is duration? 
```{r}
table(df$acquisition, df$duration == 0)
```

#Model without duration
```{r}
tree_model2 <- rpart(acquisition ~ . - duration, 
                     data = train_data, 
                     method = "class")

pred_tree2 <- predict(tree_model2, test_data, type = "class")

confusionMatrix(pred_tree2, test_data$acquisition)
```
#Drop leakage variables 
```{r}
df_clean <- df |>
  select(acquisition, acq_exp, acq_exp_sq, profit, industry, revenue, employees)
```

#Train/Test split
```{r}
set.seed(123)

df_clean$acquisition <- as.factor(df_clean$acquisition)

trainIndex <- createDataPartition(df_clean$acquisition, p = 0.8, list = FALSE)

train_clean <- df_clean[trainIndex, ]
test_clean  <- df_clean[-trainIndex, ]
```

#Cleaned Decision Tree
```{r}
tree_clean <- rpart(acquisition ~ ., 
                    data = train_clean, 
                    method = "class")

rpart.plot(tree_clean)
```

#Evaluate new model
```{r}
pred_clean <- predict(tree_clean, test_clean, type = "class")

confusionMatrix(pred_clean, test_clean$acquisition)
```

#Keep only true "pre-acquisition" variable
```{r}
df_real <- df |>
  select(acquisition, industry, revenue, employees)
```

#Re-run model
```{r}
set.seed(123)

df_real$acquisition <- as.factor(df_real$acquisition)

trainIndex <- createDataPartition(df_real$acquisition, p = 0.8, list = FALSE)

train_real <- df_real[trainIndex, ]
test_real  <- df_real[-trainIndex, ]

tree_real <- rpart(acquisition ~ ., data = train_real, method = "class")

rpart.plot(tree_real)

pred_real <- predict(tree_real, test_real, type = "class")

confusionMatrix(pred_real, test_real$acquisition)
```

PEYTON

```{r}
# Person C: Random Forest + Model Comparison 

library(tidyverse)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
```

```{r}
df <- read.csv("clean_customerretention_dataset.csv")
```

```{r}
df_real <- df |>
  select(acquisition, industry, revenue, employees)
```

```{r}
df_real$acquisition <- as.factor(df_real$acquisition)
```

# train/test

```{r}
set.seed(123)

trainIndex <- createDataPartition(df_real$acquisition, p = 0.8, list = FALSE)

train_real <- df_real[trainIndex, ]
test_real  <- df_real[-trainIndex, ]

dim(train_real)
dim(test_real)

prop.table(table(train_real$acquisition))
prop.table(table(test_real$acquisition))
```

# Log Regression

```{r}
set.seed(123)

log_model <- train(
  acquisition ~ .,
  data = train_real,
  method = "glm",
  family = "binomial")

log_pred <- predict(log_model, newdata = test_real)

log_cm <- confusionMatrix(log_pred, test_real$acquisition)

log_cm
```

# Decison Tree

```{r}
set.seed(123)

tree_real <- rpart(
  acquisition ~ .,
  data = train_real,
  method = "class")

rpart.plot(tree_real)

pred_real <- predict(tree_real, test_real, type = "class")

tree_cm <- confusionMatrix(pred_real, test_real$acquisition)

tree_cm
```

# Random Forest

```{r}
set.seed(123)

rf_model <- train(
  acquisition ~ .,
  data = train_real,
  method = "rf",
  trControl = trainControl(method = "cv", number = 10),
  tuneLength = 5)

rf_model
```
Best mtry = 2
Accuracy = 0.735
Kappa = 0.383

```{r}
rf_pred <- predict(rf_model, newdata = test_real)

rf_cm <- confusionMatrix(rf_pred, test_real$acquisition)

rf_cm
```

```{r}
rf_importance <- varImp(rf_model)

rf_importance

plot(rf_importance, top = 10,
     main = "Random Forest Variable Importance")
```

# Model Comparison

```{r}
comparison <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    log_cm$overall["Accuracy"],
    tree_cm$overall["Accuracy"],
    rf_cm$overall["Accuracy"]),
  Sensitivity = c(
    log_cm$byClass["Sensitivity"],
    tree_cm$byClass["Sensitivity"],
    rf_cm$byClass["Sensitivity"]),
  Specificity = c(
    log_cm$byClass["Specificity"],
    tree_cm$byClass["Specificity"],
    rf_cm$byClass["Specificity"]))

comparison
```

